An Analysis of Health & Lifestyle Factors Associated with Sleep Quality

### 2024-04-18 products/manuscript/Manuscript.qmd
Added Discussion section and edited manuscript
Wrote discussion section. Improved results section with more figures, descriptions of model creation, and descriptions of results. Edited writing in other parts

---
title: "An Analysis of Health & Lifestyle Factors Associated with Sleep Quality"
subtitle: ""
author: Kevin Kosewick and Emma Hardin-Parker
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../manuscript/dataanalysis-references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Additionally, based on the findings of Sun et. al, that obesity can negatively impact sleep quality, we also expect that higher levels of blood pressure will result in poorer sleep quality. The American Sleep Foundation states that both short and long sleep duration increase risk of hypertension in adults (Li & Shang, 2021). This draws interesting focus to the circular nature of sleep quality and hypertension. Those with hypertension also face a higher risk of insomnia, a disorder affecting the ability to fall asleep, maintain sleep, and sleep without continuous disturbance (Liu et. al, 2022). Short sleep duration is also positively associated with stress level (APA, 2013). Survey findings indicate that the average American sleeps 6.7 hours per night, which is well below the minimum recommended duration seven to nine hours (APA, 2013). Similar to the relationship between hypertension and sleep quality, as stress level increases, sleep duration decreases thereby causing poorer sleep quality. Poorer sleep quality then leads to higher stress, greater risk of hypertension, and shorter sleep duration.

@@

The dataset originally contained 373 observations for 13 variables
covering a broad spectrum health and lifestyle variables associated with
sleep quality. The curators of the dataset did not provide information
on how or where this data was collected, so we solely relied on the
codebook for variable definitions. There were no missing values or
erroneous variables in the dataset that we needed to remove. The
"Blood.Pressure" variable was transformed to reflect the individual systolic and
diastolic blood pressure measurements from each subject. We then chose
to add an additional variable, labeled 'cat_bp' to reflect the
categorical blood pressure status of a subject based on the [American
Heart Association
guidelines](https://www.heart.org/en/health-topics/high-blood-pressure).
Additionally we created a variable, 'StepsGroup' to determine the
categorical activity level of subjects in the dataset based on these
guidelines provided by [10,000 steps.org] (https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500)
Finally, we created an additional categorical variable,
'PhysicalActivityGroup' based on 30 minute increments, to categorically
represent the level of physical activity reported by a subject. The
outcome of interest is noted as "Quality.of.Sleep" that we will
reference as Sleep Quality or Quality of Sleep throughout this report.

@@

Given that there were only 13 original variables, and 17 variables after
final data transformation; simple linear regression models were
performed to determine baseline associations between variables and Sleep
Quality, the outcome of interest. These regression models were created
to determine which variables had the largest impact on Sleep Quality and
drove the rest of the analyses. Simple linear regression models were
fitted for the following variables: BMI, cat_bp, Stress Level, Gender,
Age, and Occupation.

@@

The objective of our analysis was to determine which of these variables
are most important for determining an individual's sleep quality and
what specific effect they have. After conducting an (EDA) we
hypothesized that stress, sleep duration, and occupation would have the
largest impact on sleep quality. We believed that stress and sleep
quality would be negatively correlated, sleep duration and quality would
be positively correlated, and that jobs with longer or more variable
hours would negatively affect sleep quality.

@@

Random Forest Models and Cross-Validation were considered to be the
"best" fit for our outcome of interest, Quality of Sleep. The data were
not split into train/test subsets as the data contains less than 400
observations and many of the values were unique in comparison to the
rest of the data. Prior to constructing this model, a collinearity plot
was constructed to determine any presence of collinearity in our data.
![Collinearity between all variables.](images/collinearity.png)

@@

A variable importance plot (VIP) was constructed to determine which
variables had the strongest impact in our Random Forest Model. ![The
ranking of each variable\'s importance in determining sleep quality
according to our Random Forest
model.](images/variableimportance.png)Sleep Duration, Stress Level, and
Age had the highest levels of importance in our model. Physical Activity
Level was excluded entirely, which was to be expected given our
exploratory data analysis and simple linear regression models.

@@

*DEFINE DECISION TREES. DEFINE BAGGING AND BOOTSTRAPPING. DEFINE RANDOM
FOREST MODELS* *DEFINE CROSS-VALIDATION* *DEFINE IMPORTANCE OF NULL
MODEL TO COMPARE RESULTS*

*INCLUDE CITATIONS OF CHAPTER 9,10,11 OF HMLR LINKED IN "MANY TREE
MODELS" SECTION OF CONTENT FOLDER 11*

@@

We used both RMSE and R² to evaluate the performance of our model
compared to a "null model". Null models generate predictions for an
outcome of interest without using any predictors. Instead, they simply
average the values of the outcome of interest to make "predictions".

![The results generated from our RF model. RMSE increases when more
predictors are added to the model. The R-squared value also decreases as
more predictors are added to the model.](images/RFtuningresults.png)

Figure (*name/order figures after we finalize draft*) shows that as we
include more predictors in our model, RMSE increases slightly while the
R-squared value decreases. This indicates that some variables in our
dataset are more important for predicting sleep quality than the others.
Based off of our variable importance graph, we can assume that stress
level, sleep duration, and age are the factors that carry the most
weight when predicting sleep quality.

We went on to evaluate how well our RF model predicted values by
creating an observed vs. predicted plot.

![A comparison of the actual values for sleep quality in our dataset
(observed) compared to the values predicted by our fitted RF
model.](images/observed_vs_predicted.png)

@@

This analysis was conducted under R version 4.3.1 on a MacOS operating
system. The following R packages were used in the development
of these analyses:here, skimr, broom, tidyverse,
ggplot2, dplyr, corrplot, ranger, and vip. All processing and analysis code
can be found in the Supplementary Material file.

@@

The 'Quality of Sleep' variable has been selected as our outcome of
interest. It is important to note that this variable is a subjective
measure of a participant's self-reported sleep quality. The following
plot shows that most subjects reported a sleep quality score of 8 (out
of 10).

![Distribution of Sleep Quality Scores in the
dataset.](images/unnamed-chunk-17-1.png)

## Exploratory Findings

@@

*To Emma: did you want this text in both places you placed it?*
*To Kevin: I placed in both sections as a placeholder, whichever you think is best!*

A
variable importance plot (VIP) was constructed to determine which
variables had the strongest impact in our Random Forest Model. *INCLUDE
PLOT HERE* Sleep Duration, Stress Level, and Age had the highest levels
of importance in our model. Physical Activity Level was excluded
entirely, which was to be expected given our exploratory data analysis
and simple linear regression models.

*Include comparison of RF Model and null model*

@@

*PART 5*

@@

*PART 5*

@@

*PART 5*

*Include that all parameters were self-reported and thus subjected to
bias*

@@

*PART 5*

@@

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9815923/#:~:text=Patients%20with%20hypertension%20commonly%20complain,an%20increased%20risk%20of%20insomnia. --\> hypertension and insomnia

@@

https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500

>>>>

---
title: "An Analysis of Health & Lifestyle Factors Associated with Sleep Quality"
subtitle: ""
author: Kevin Kosewick and Emma Hardin-Parker
date: "4-18-2024"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../manuscript/dataanalysis-references.bib
editor: 
  markdown: 
    wrap: 72
---

@@

Additionally, based on the findings of Sun et. al that obesity can
negatively impact sleep quality, we also expect that higher levels of
blood pressure will result in poorer sleep quality. The American Sleep
Foundation states that both short and long sleep duration increase risk
of hypertension in adults (Li & Shang, 2021). This draws interesting
focus to the circular nature of sleep quality and hypertension. Those
with hypertension also face a higher risk of insomnia, a disorder
affecting the ability to fall asleep, maintain sleep, and sleep without
continuous disturbance (Liu et. al, 2022). Short sleep duration is also
positively associated with stress level (APA, 2013). Survey findings
indicate that the average American sleeps 6.7 hours per night, which is
well below the minimum recommended duration seven to nine hours (APA,
2013). Similar to the relationship between hypertension and sleep
quality, as stress level increases, sleep duration decreases thereby
causing poorer sleep quality. Poorer sleep quality then leads to higher
stress, greater risk of hypertension, and shorter sleep duration.

@@

The dataset originally contained 373 observations for 13 variables
covering a broad spectrum of health and lifestyle variables associated
with sleep quality. The curators of the dataset did not provide
information on how or where this data was collected, so we solely relied
on the codebook for variable definitions. There were no missing values
or erroneous variables in the dataset that we needed to remove. The
"Blood.Pressure" variable was transformed to reflect the individual
systolic and diastolic blood pressure measurements from each subject. We
then chose to add an additional variable, labeled 'cat_bp' to reflect
the categorical blood pressure status of a subject based on the
[American Heart Association
guidelines](https://www.heart.org/en/health-topics/high-blood-pressure).
Additionally we created a variable named 'StepsGroup' to determine the
categorical activity level of subjects in the dataset based on these
guidelines provided by \[10,000 steps.org\]
(https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:\~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500)
Finally, we created an additional categorical variable called
'PhysicalActivityGroup' to categorically represent the level of physical
activity reported by a subject. The levels are differentiated by 30
minute intervals of weekly physical activity. The outcome of interest is
noted as "Quality.of.Sleep" that we will reference as Sleep Quality or
Quality of Sleep throughout this report.

@@

Given that there were only 17 variables after final data transformation,
simple linear regression models were performed to determine baseline
associations between variables and Sleep Quality (the outcome of
interest). These regression models were created to determine which
variables had the largest impact on Sleep Quality and drove the rest of
the analyses. Simple linear regression models were fitted for the
following variables: BMI, cat_bp, Stress Level, Gender, Age, and
Occupation.

@@

The objective of our analysis was to determine which of these variables
are most important for determining an individual's sleep quality and
what specific effect they have. After conducting an EDA we hypothesized
that stress, sleep duration, and occupation would have the largest
impact on sleep quality. We believed that stress and sleep quality would
be negatively correlated, sleep duration and quality would be positively
correlated, and that jobs with longer or more variable hours would
negatively affect sleep quality.

@@

Random Forest Models and 5-fold Cross-Validation were considered to be
the "best" fit for our outcome of interest, Quality of Sleep. The data
were not split into train/test subsets as the data contains less than
400 observations and many of the values were unique in comparison to the
rest of the data. Prior to constructing this model, a colinearity plot
was constructed to determine any presence of colinearity in our data.

![Figure 1: A colinearity plot demonstrating the colinearity between our
variables.](images/collinearity.png){width="586"}

@@

@@

We fit GLMs for each predictor in order to determine their effect on
sleep quality. Each model only used a single predictor as the only aim
was to quantify and understand the impact of each for our outcome of
interest. For our Random Forest model, we used mtry (with a range of
1-7) and min_n (with a range of 1-21) as tuning parameters to create 300
different decision trees. Mtry controls how many predictors a decision
tree can consider at any given point in time. It adds randomness to the
decision tree creation process and ensures that all of the trees do not
look the same. Min_n specifies the minimum number of samples that should
be present in the leaf node after splitting a node. In our case, this
means that if any terminal node has more than 21 observations and is not
a pure node, we can split it further into subnodes. This was all done
using 5-fold Cross-Validation repeated 5 times in order to train the
data to predict on an unseen dataset.

@@

Null models provide a baseline for the evaluation of a model's
performance. They generate predictions for an outcome of interest
without using any predictor variables. Instead, they simply average the
values of the outcome of interest to make these "predictions". We used
both RMSE and R² values to evaluate the performance of our model
compared to a "null model".

@@

This analysis was conducted under R version 4.3.1 on a MacOS operating
system. The following R packages were used in the development of these
analyses: here, skimr, broom, tidyverse, ggplot2, dplyr, corrplot,
ranger, and vip. All processing and analysis code can be found in the
Supplementary Material file.

@@

The 'Quality of Sleep' variable has been selected as our outcome of
interest. It is important to note that this variable is a subjective
measure of a participant's self-reported sleep quality. Figure 2 shows
that most subjects reported a sleep quality score of 8 (out of 10).

![Figure 2: Distribution of Sleep Quality Scores in the
dataset.](images/unnamed-chunk-17-1.png)

Our Random Forest Model built with 5-fold Cross-Validation explored
different combinations of predictors to determine which parameters gave
the most accurate predictions of sleep quality.

![Figure 3: The results generated from our RF model. RMSE increases when
more predictors are added to the model. The R-squared value also
decreases as more predictors are added to the
model.](images/RFtuningresults.png)

Figure 3 shows that as we include more predictors in our model, RMSE
increases slightly while the R-squared value decreases. This indicates
that some variables in our dataset are more important for predicting
sleep quality than the others. Based off of our variable importance
graph, we can assume that stress level, sleep duration, and age are the
factors that carry the most weight when predicting sleep quality.

We went on to evaluate how well our RF model predicted values by
creating an observed vs. predicted plot as seen in figure 4.

![Figure 4: A comparison of the actual values for sleep quality in our
dataset (observed) compared to the values predicted by our fitted RF
model.](images/observed_vs_predicted.png)

![Figure 5: The observed vs. predicted values for our "Null
model".](images/null_observedvspredicted.png)

We can see in figure 5 that the null model only predicts values that are
the average of our observed values. As expected, these predictions are
far from our actual observed values.

We believed that stress and sleep quality would be negatively
correlated, sleep duration and quality would be positively correlated,
and that jobs with longer or more variable hours would negatively affect
sleep quality. We found that all of our assumptions were supported; the
results of our generalized linear models (GLMs) are included in the
supplementary material folder as there are too many of them to include
in the main manuscript.

@@

A variable importance plot (VIP) (figure 6) was constructed using the
"best" set of parameters (as guided by our RF model) to determine which
variables had the strongest impact on sleep quality.

![The ranking of each variable's importance in determining sleep quality
according to our Random Forest model.](images/variableimportance.png)

As seen in figure 6, Sleep Duration, Stress Level, and Age had the
highest levels of importance in our model. Physical Activity Level was
excluded entirely, which was to be expected given our exploratory data
analysis and simple linear regression models.

![Table 1: A comparison of the RMSE and R-squared values generated by
the null and RF models. The null model has no R-squared value due to its
method of generating predictions.](images/metric_table.png){width="510"}

As seen in table 1, our null model was found to have an RMSE of 1.193;
compared to our best parameter RF model's RMSE of 0.059, this is
relatively high. The R-squared value doesn't exist for the null model
since there is technically no variation between the observed and
predicted values.

@@

@@

The differences between the null model and RF model's metrics show that
our RF model is better suited for predicting sleep quality on unseen
data than a model using no predictors. The R-squared value of 0.997 is
extremely high; this shows that a large amount of variation is explained
using our RF model. The variable importance plot we created answers our
questions about which variables are most important for predicting sleep
quality. As we can see, sleep duration, stress level, and age were the
top three predictors for sleep quality in our data. This reflects the
relationships we saw in our colinearity plot as well as the clear
correlations we found between the variables in our EDA. We did not
expect to see age so high on the list; this could be due to us grouping
ages together in our EDA and therefore misleading us in our preliminary
data exploration. We did not use the groups when creating our models as
we learned that grouping continuous variables is a poor practice as we
progressed in our MADA course.

Through our linear models (included in the supplementary materials
folder), we determined that our hypotheses regarding each variable's
effect on quality of sleep were supported. The variables determined to
be the most important by our RF model had the following effects on sleep
quality: sleep duration was positively correlated, age was positively
correlated, and stress level was negatively correlated.

@@

The main limitations of our models stemmed from the data itself. As
stated in our introduction section, we have been unable to identify
exactly where the data were collected and how it was gathered. The
values for stress level and sleep duration were also based on a
subjective rating given by each participant in the study; this could
introduce individual bias to the data and prevent an objective
evaluation of the effect of these variables. An inherent limitation in
RF models is difficulty of interpretation; we attempted to minimize this
by creating easy to read tables and graphs for the results. Future
analyses would benefit from more detailed meta data and a better
understanding of the experimental design that generated our
observations.

The strengths of our analysis lie in the easily interpretation of our
GLMs and the predictive power of an RF model. We were able to clearly
display and quantify the relationships between our predictors and sleep
quality, answering our question about each variable's effect on our
outcome of interest. We were also able to create a powerful RF model
that explained nearly all of the variation in our data as well as a
ranking of each variable's impact on sleep quality. Fitting our RF model
with CV allowed us to train the model to create predictions for "unseen"
data as well.

@@

Overall, while our study limitations prevent us from generalizing the
findings of this analysis broadly, we were able to clearly answer our
questions about the data. Our findings supported our hypotheses about
the effects of each predictor on sleep quality. However, our hypothesis
about which variables would be the most important was not supported by
the analysis. Instead, we determined that stress level, sleep duration,
and age were the most important predictors for our outcome of interest.
Ultimately, the results indicate that the sleep quality of this study's
participants is impacted disproportionately by an assortment of
variables dealing with physical, mental, and environmental factors.

@@

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9815923/#:\~:text=Patients%20with%20hypertension%20commonly%20complain,an%20increased%20risk%20of%20insomnia.
--\> hypertension and insomnia

@@

https://www.10000steps.org.au/articles/healthy-lifestyles/counting-steps/#:\~:text=The%20following%20pedometer%20indices%20have%20been%20developed%20to,day%205%20Highly%20active%20is%20more%20than%2012%2C500
